
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This lexicon has been modified from the following template:
%
% Dictionary
% LaTeX Template
% Version 1.0 (20/12/14)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Vel (vel@latextemplates.com) inspired by a template by Marc Lavaud
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\fancyhead[RO]{\textsf{\rightmark -- \leftmark}} % Top left header
\fancyhead[LO]{\textsf{\thepage}} % Top right header

\fancyhead[LE]{\textsf{\rightmark -- \leftmark}} % Top left header
\fancyhead[RE]{\textsf{\thepage}} % Top right header

%default entry command
%\newcommand{\entry}[4]{\markboth{#1}{#1}\textbf{#1}\ {(#2)}\ \textit{#3}\
  %$\bullet$\ {#4}}  % Defines the command to print each word on the page,
                    % \markboth{}{} prints the first word on the page in the top
                    % left header and the last word in the top right

\newcommand{\entry}[4]{\markboth{#1}{#1}\textbf{#1}\ {[\textipa{#2}]}\ \textit{#3}\
  \ {#4}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% LEXICON
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{A}

\begin{multicols*}{2}

\entry{kujari}{kuZa\textturnrrtail i}{na}{Southern Cassowary; \textit{(fig)} a
  person prone to listlessness}

\begin{remark}
The Southern Cassowary is seen as untamable and always in control. However, it
seems to have no definite plans and spends its time wandering aimlessly. This
sentiment can be applied to people.
\end{remark}

\entry{wurr}{wu\|]R}{v}{(\textbf{0}) it is electrically storming, there is
  lightning; (1) to be struck by lightning; (2) to strike}

\entry{Abbreviated}{uh-bree-vee-ey-tid}{Adjective}{Constituting a shorter or
  smaller version of}

\entry{Aberration}{ab-uh-rey-shuh n}{Noun}{The act of deviating from the
  ordinary, usual, or normal type.}

\entry{Above}{uh-buhv}{Preposition}{In extended space over and not touching.}

\entry{Academia}{ak-uh-dee-mee-uh}{Noun}{The environment or community concerned
  with the pursuit of research, education, and scholarship.}

\entry{Accomplished}{uh-kom-plisht}{Adjective}{Completed; done; effected. Highly trained or skilled in a particular activity.}

\entry{Algebra}{al-juh-bruh}{Noun}{The part of mathematics in which letters and other general symbols are used to represent numbers and quantities in formulae and equations.}

\entry{Amatol}{am-uh-tawl}{Noun}{A high explosive consisting of a mixture of TNT and ammonium nitrate.}

\entry{Almanac}{awl-muh-nak}{Noun}{An annual publication containing a calendar for the coming year, the times of such events and phenomena}

\entry{Animal}{an-uh-muh l}{Noun}{A living organism which feeds on organic matter, typically having specialized sense organs and nervous system and able to respond rapidly to stimuli.}

\entry{Ascension}{auh-sen-shuh n}{Noun}{The action of rising to an important position or a higher level.}

\entry{Aspire}{uh-spahyuh r}{Verb}{Direct one's hopes or ambitions towards achieving something.}

\entry{Athlete}{ath-leet}{Noun}{a person trained or gifted in exercises or contests involving physical agility, stamina, or strength; a participant in a sport, exercise, or game requiring physical skill.}

\entry{Azobenzene}{az-oh-ben-zeen}{Noun}{A synthetic crystalline organic compound used chiefly in dye manufacture.}

\entry{something}{az-oh-ben-zeen}{Noun}{A synthetic crystalline organic compound used chiefly in dye manufacture.}

\entry{Babble}{bab-uh l}{Verb}{Talk rapidly and continuously in a foolish, excited, or incomprehensible way.}

\entry{Balance}{bal-uh ns}{Noun}{An even distribution of weight enabling someone or something to remain upright and steady. An instrument for determining weight, typically by the equilibrium of a bar with a fulcrum at the center, from each end of which is suspended a scale or pan, one holding an object of known weight, and the other holding the object to be weighed.}

\entry{Barbet}{bahr-bit}{Noun}{A large-headed, brightly coloured fruit-eating bird that has a stout bill with tufts of bristles at the base. Barbets are found on all continents, especially in the tropics.}

\entry{Beetroot}{beet-root}{Noun}{The edible dark-red spherical root of a kind of beet, eaten as a vegetable.}

\entry{Besides}{bih-sahydz}{Preposition}{In addition to; apart from.}

\entry{Bevel}{bev-uh l}{Noun}{A slope from the horizontal or vertical in carpentry and stonework; a sloping surface or edge.}

\entry{Bevel}{bev-uh l}{Noun}{A slope from the horizontal or vertical in carpentry and stonework; a sloping surface or edge.}

\entry{Biennial}{bahy-en-ee-uh l}{Adjective}{Taking place every other year.}

\entry{Bioinformatics}{bahy-oh-in-fer-mat-iks}{Noun}{The retrieval and analysis of biochemical and biological data using mathematics and computer science, as in the study of genomes. The science of collecting and analysing complex biological data such as genetic codes.}

\entry{Bleep}{bleep}{Noun}{A short high-pitched sound made by an electronic device as a signal or to attract attention.}

\entry{Blind}{blahynd}{Adjective}{Unable to see; lacking the sense of sight; sightless.}

\entry{Bonanza}{buh-nan-zuh}{Noun}{A situation which creates a sudden increase in wealth, good fortune, or profits.}

\entry{Book}{boo k}{Noun}{A written or printed work consisting of pages glued or sewn together along one side and bound in covers.}

\entry{Bran}{bran}{Noun}{The partly ground husk of wheat or other grain, separated from flour meal by sifting.}

\entry{Break}{breyk}{Verb}{Separate into pieces as a result of a blow, shock, or strain.}

\entry{Bridge}{brij}{Noun}{A structure carrying a road, path, railway, etc. across a river, road, or other obstacle. Music: The part of a stringed instrument over which the strings are stretched.}

\entry{Brioche}{bree-ohsh}{Noun}{A light sweet yeast bread typically in the form of a small round roll.}

\entry{Buzzard}{buhz-erd}{Noun}{A large hawklike bird of prey with broad wings and a rounded tail, often seen soaring in wide circles.}

\entry{Bystander}{bahy-stan-der}{Noun}{A person who is present at an event or
  incident but does not take part.}

\end{multicols*}
